---
title: "Analysis of Cyclistics' casual and member users"
author: "Antonio Colantuono"
date: "2022-12-20"
output: 
  html_document:
    toc: true
  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Libraries and locale

Import the libraries that are going to be needed for this analysis:

```{r libraries, warning=FALSE, message=FALSE, error=FALSE}
library(tidyverse)
library(dplyr)
library(tidyr)
library(lubridate)
```

Set the locale to an english speaking country so that months() and weekdays() function return values in english:

```{r locale, warning=FALSE, message=FALSE, error=FALSE}
Sys.setlocale("LC_TIME","uk")
```

## Data loading

Load all the twelve datasets and create a new dataset containing all the data: 

```{r datasets, warning=FALSE, message=FALSE, error=FALSE}
data1 <- read.csv("202201-divvy-tripdata.csv")
data2 <- read.csv("202202-divvy-tripdata.csv")
data3 <- read.csv("202203-divvy-tripdata.csv")
data4 <- read.csv("202204-divvy-tripdata.csv")
data5 <- read.csv("202205-divvy-tripdata.csv")
data6 <- read.csv("202206-divvy-tripdata.csv")
data7 <- read.csv("202207-divvy-tripdata.csv")
data8 <- read.csv("202208-divvy-tripdata.csv")
data9 <- read.csv("202209-divvy-tripdata.csv")
data10 <- read.csv("202210-divvy-tripdata.csv")
data11 <- read.csv("202211-divvy-tripdata.csv")
data12 <- read.csv("202112-divvy-tripdata.csv")

data <- bind_rows(data1, data2, data3, data4, data5, data6, data7, data8, data9,
                  data10, data11, data12)
head(data)
```

## Data cleaning

``` {r clean, warning=FALSE, message=FALSE, error=FALSE}
potd <- c("Night", "Morning", "Afternoon", "Evening")

cleaned_data <- data %>%
  mutate(travel_time = int_length(interval(started_at, ended_at))) %>%
  mutate(weekday = weekdays(as.Date(started_at))) %>%
  mutate(month = months(as.Date(started_at))) %>%
  mutate(part_of_the_day = potd[hour(started_at)%/%6 + 1]) %>%
  rename(user_type = member_casual) %>%
  select(ride_id, rideable_type, user_type, travel_time, part_of_the_day, weekday, month) %>%
  filter(travel_time>0)

cleaned_data$weekday <- factor(cleaned_data$weekday, 
                               levels = c("Monday", "Tuesday", "Wednesday", 
                                          "Thursday", "Friday", "Saturday", "Sunday"))

cleaned_data$part_of_the_day <- factor(cleaned_data$part_of_the_day, 
                                       levels = c("Night", "Morning", 
                                                  "Afternoon", "Evening"))
cleaned_data$month <- factor(cleaned_data$month, 
                             levels = c("January", "February", "March", "April",
                                        "May", "June", "July", "August", "September",
                                        "October", "November","December"))
head(cleaned_data)
```

Generate dataframes containing aggregated data for several purposes:

``` {r aggregate, warning=FALSE, message=FALSE, error=FALSE}
extract_daily <- cleaned_data %>%
  group_by(user_type, part_of_the_day) %>%
  summarise(amount_of_rides = n(), average_length = mean(travel_time)/60)

extract_weekly <- cleaned_data %>%
  group_by(user_type, weekday) %>%
  summarise(amount_of_rides = n(), average_length = mean(travel_time)/60)

extract_monthly <- cleaned_data %>%
  group_by(user_type, month) %>%
  summarise(amount_of_rides = n(), average_length = mean(travel_time)/60)
```

## Charts creation

### Daily

``` {r daily}

ggplot(data=extract_daily, mapping = aes(x=part_of_the_day, y=amount_of_rides, 
                                        color=user_type, group=user_type)) + geom_point() + geom_line() +
                                        theme(axis.text.x = element_text(angle=45, vjust = 0.75))

ggplot(data=extract_daily, mapping = aes(x=part_of_the_day, y=average_length, 
                                        color=user_type, group=user_type)) + geom_point() + geom_line() +
                                        theme(axis.text.x = element_text(angle=45, vjust = 0.75))
```

### Weekly
```{r weekly}
ggplot(data=extract_weekly, mapping = aes(x=weekday, y=amount_of_rides, 
                                        color=user_type, group=user_type)) + geom_point() + geom_line() +
                                        theme(axis.text.x = element_text(angle=45, vjust = 0.75))

ggplot(data=extract_weekly, mapping = aes(x=weekday, y=average_length, 
                                       color=user_type, group=user_type)) + geom_point() + geom_line() +
                                       theme(axis.text.x = element_text(angle=45, vjust = 0.75))
```

### Monthly
```{r monthly}
ggplot(data=extract_monthly, mapping = aes(x=month, y=amount_of_rides, 
                                        color=user_type, group=user_type)) + geom_point() + geom_line() +
                                        theme(axis.text.x = element_text(angle=45, vjust = 0.75))

ggplot(data=extract_monthly, mapping = aes(x=month, y=average_length, 
                                         color=user_type, group=user_type)) + geom_point() + geom_line() +
                                         theme(axis.text.x = element_text(angle=45, vjust = 0.75))
```











